ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-06-01 paper/older-format-version/Manuscript-Data-Package.Rmd
Create a document with all manuscript changes and add answer letters for the reviewers.

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an R data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) of Canada's General Social Survey. Active travel is characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting."

@@

{ActiveCA} is distributed as an R package with a number of data 
objects and their documentation. R packages contain code, data, and documentation
in a standardized format that can be installed by R users via a software repository,
such as CRAN (Comprehensive R Archive Network) or GitHub, which makes them an 
adroit medium to distribute analysis-ready data.

@@

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29). Each row in the table correspond to a survey respondent, while the columns refer the following information: record identification (PUMFID), the person's weight (WGHT_PER), the month the survey data was collected (SURVMNTH), the respondent's age group (AGEGR10), the respondent's sex (SEX), and the respondent's marital status (MARSTAT).

The Main File of the 2015 GSS surveys includes a total of 17,390 respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and 848 variables.  For discrete variables, Statistics Canada has assigned specific codes to the possible values, with each code accompanied by a label. For instance, in the case of the variable SURVMNTH, a value of 1 means January 2016, 2 means February 2016, 3 corresponds to March 2016, and so on.

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an example of the Main File structure, Table \ref{tab:ep-2015-unprocessed} presents the first seven episodes for the record identification number 10041 and some variables from the GSS PUMF 2015 Episode File (Cycle 29). Each row in the table corresponds to an episode associated with the specified record identification (PUMFID = 10041), including the episode's weight (WGHT_EPI), episode number (EPINO), activity code (TUI_01), episode duration (DURATION), and episode location (LOCATION).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this respondent began the diary description by sleeping at home (TUI_01 = 1 and LOCATION = 300) for 210 minutes, followed by 40 minutes of personal hygiene (TUI_01 = 2). The respondent then spent 15 minutes on personal care activities, such as getting ready for school, supervising homework, reading, playing, reprimanding, or providing educational or emotional support, as indicated by TUI_01 = 27. Next, they recorded a travel episode, walking for 15 minutes (TUI_01 = 7 and LOCATION = 315), where both the origin and destination were their home. Such trips, where the journey starts and finishes at home, can be classified as recreational or leisure trips. Next, the respondent spent 3 hours searching for a job (TUI_01 = 9), took a 15-minute lunch break (TUI_01 = 6), and then cleaned the house (TUI_01 = 18) for two hours. Table \ref{tab:ep-2015-unprocessed} displays only six variables out of the 527 available. As shown, since the dataset does not label the variable values, decoding them can be both time-consuming and challenging.

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) |>
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number 10041.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: PUMFID: record identification. EPINO: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

The documentation of the package can be accessed by running ?ActiveCA in R. The documentation includes a description of the package, information about the authors, contributors, and maintainers, and an overview of all pre-processed data sets included in the package.

Figure \ref{fig:walking_documentation} displays the documentation for the walking episodes from the 2015 GSS (walking_2015.rda). The documentation provides a brief description of the dataset, instructions on how to load the package, as well as details about the file format and descriptions of all variables (already labeled).

@@

For instance, to calculate the percentage of respondents from the 2015 GSS survey with active travel episodes, it is necessary to account for the person's weight. In 2015, the weight variable is represented by WGHT_PER. The code below demonstrates how to obtain the percentage of people with active travel episodes by age group. It uses the dplyr package [@dplyr2023] to manipulate the data.

The process begins by creating a dataset that sums the population by age group. Then, it joins the 2015 episodes with the 2015 Main File. Note that in both operations, the code sums the person weight variable (WGHT_PER) to obtain the correct population values. After this, a new dataset, Active_percentage, is created by merging both previous datasets. The percentage of active travel episodes by age group is calculated by dividing the total population by the population with active trip episodes, then multiplying by 100 and rounding with 2 decimal places.

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29), displayed in \ref{tab:main-2015-unprocessed} before our processing. Table \ref{tab:ep-2015-processed} presents the walking episodes for the record identification number 10041 from the GSS PUMF 2015 Episode File (Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only the unique active travel episode appear in Table \ref{tab:ep-2015-processed} since the records were filtered to select cases with walking or cycling episodes. For both cases, Tables \ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain labeled variables, facilitating the interpretation of the data.

@@

Table \ref{tab:table-02} and \ref{tab:table-03} provide descriptive
statistics for the two modes of transportation, split by destination
categories, from 1986 to 1998 and from 2005 to 2015, respectively. In
Table \ref{tab:table-02}, one can observed that in 1986 and 1992,
walking trips destined for home had the highest medians. However, by
1998, the highest medians shifted to trips to work or school, a
transition that also occurred for cycling trips between 1992 and 1998.
Table \ref{tab:table-03} indicates that the median duration for walking trips to
home and work or school remained at 10 minutes.

@@

In 1992, walking trips with home as both the origin and destination
made up the majority, accounting for about 30% of all walking trips.
These trips often involved leisure activities, like short walks or dog
walking. Following this, trips from home to work or school comprised
18% of walking trips. Overall, home emerged as a crucial hub, either
as an origin or destination, with only 5% of trips not involving home.
By 2015, home remained a significant node, but new locations
distributed the proportion of trips to areas not considered in 1992. In
2015, the highest proportion of trips were from home to
work or school (12%) and vice versa (11%). home to home accounted
for 8% of trips, and grocery stores became a notable destination for
those leaving home (6%), surpassing trips to other's home (4%).

@@

For cycling trips, Figure \ref{fig:figure-02}, shows that in 1992, when
this mode of transportation was first included as an activity, the
majority of trips were from home to work or school, accounting for
about 25% of cases. This pattern remained in 2015, with these trips
representing 30% of the cases. However, a notable change occurred in
home to home trips, which decreased significantly from 19% in 1992
to 5% in 2015.

@@

This paper presents {ActiveCA}, an open data product that provides analysis-ready 
data from Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) 
of the GSS surveys on active travel in Canada. In the form of an R data package, 
{ActiveCA} was developed after collecting, cleaning, and processing the survey data, 
providing information on origins, destinations, and duration of active travel, as
well other information.

It is important to remark that the present version of {ActiveCA} covers all Canadian
time use surveys up to 2015. While the most recent time use survey was carried out
in 2022 [@wray2024], the Public Use Microdata Files are currently unavailable, and
at the moment it is estimated that they will only be published in the later part
of 2025. The R package will be updated once these files become available.

The value of {ActiveCA} lies in its transparency, accessibility, and ease of use,
which facilitates the addition of complementary data sets in the future. R users
can seamlessly explore GSS walking and cycling episodes, 
with the option to suggest enhancements to the package as needed. This article adopts
the structure proposed by Anastasia and Páez [-@soukhov2023], whose work provided
essential guidance for the creation of this package. Similarly, we aim to contribute
to the academic community by promoting transparent research practices that encourage
replication and innovation in related fields. We believe that {ActiveCA} will serve
as a basis for further research on GSS and for the integration of additional data by
the authors or the wider open source community.

>>>>

---
title: "ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel"
author:
  - name: "Anon1"
    email: "Anon1@anon.org"
    affiliation: CODE
    footnote: 1
  - name: "Anon2"
    email: "Anon2@anon.org"
    affiliation: CODE
  - name: "Anon3"
    email: "Anon3@anon.org"
    affiliation: CODE
address:
  - code: CODE
    address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes {ActiveCA}, an open data product with Canadian time use data. {ActiveCA} is an `R` data package that contains analysis-ready data related to active travel spanning almost 40 years, extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) of Canada's General Social Survey. Active travel is characterized by mode, with walking being part of every cycle and bicycling starting in 1992. The attributes of active trips are the types of locations of origins and destinations, the duration of trips, and episode weights for expanding the trips to population-wide estimates. Based on the year of the survey, a variety of locations are coded. In earlier cycles, these include home, work or school, and other's home, whereas in later cycles these are augmented with locations such as grocery stores, restaurants, outdoor destinations, and others. The geographical resolution includes the province and whether the episode was in an urban or rural setting."

@@

{ActiveCA} is distributed as an `R` package with a number of data 
objects and their documentation. `R` packages contain code, data, and documentation
in a standardized format that can be installed by `R` users via a software repository,
such as CRAN (Comprehensive R Archive Network) or GitHub, which makes them an 
adroit medium to distribute analysis-ready data.

@@

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29). Each row in the table correspond to a survey respondent, while the columns refer the following information: record identification (`PUMFID`), the person's weight (`WGHT_PER`), the month the survey data was collected (`SURVMNTH`), the respondent's age group (`AGEGR10`), the respondent's sex (`SEX`), and the respondent's marital status (`MARSTAT`).

The Main File of the 2015 GSS surveys includes a total of 17,390 respondents, representing `r sum(gss_m_2015$WGHT_PER)` individuals and 848 variables.  For discrete variables, Statistics Canada has assigned specific codes to the possible values, with each code accompanied by a label. For instance, in the case of the variable `SURVMNTH`, a value of `1` means `January 2016`, 2 means `February 2016`, 3 corresponds to `March 2016`, and so on.

@@

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an example of the Main File structure, Table \ref{tab:ep-2015-unprocessed} presents the first seven episodes for the record identification number `10041` and some variables from the GSS PUMF 2015 Episode File (Cycle 29). Each row in the table corresponds to an episode associated with the specified record identification (`PUMFID = 10041`), including the episode's weight (`WGHT_EPI`), episode number (`EPINO`), activity code (`TUI_01`), episode duration (`DURATION`), and episode location (`LOCATION`).

@@

In the case illustrated in Table \ref{tab:ep-2015-unprocessed}, this respondent began the diary description by sleeping at home (`TUI_01 = 1` and `LOCATION = 300`) for 210 minutes, followed by 40 minutes of personal hygiene (`TUI_01 = 2`). The respondent then spent 15 minutes on personal care activities, such as getting ready for school, supervising homework, reading, playing, reprimanding, or providing educational or emotional support, as indicated by `TUI_01 = 27`. Next, they recorded a travel episode, walking for 15 minutes (`TUI_01 = 7` and `LOCATION = 315`), where both the origin and destination were their home. Such trips, where the journey starts and finishes at home, can be classified as recreational or leisure trips. Next, the respondent spent 3 hours searching for a job (`TUI_01 = 9`), took a 15-minute lunch break (`TUI_01 = 6`), and then cleaned the house (`TUI_01 = 18`) for two hours. Table \ref{tab:ep-2015-unprocessed} displays only six variables out of the 527 available. As shown, since the dataset does not label the variable values, decoding them can be both time-consuming and challenging.

```{r gss-epi-file-2015}
ep_2015_unprocessed <- gss_e_2015 |>
  filter(PUMFID == 10041 & EPINO <= 7) |>
  dplyr::select(PUMFID, EPINO,WGHT_EPI,TUI_01, DURATION, LOCATION) |>
  kable(
    format = "latex", 
    booktabs = TRUE, 
    longtable = TRUE, 
    align = c(rep("c", ncol(gss_m_2015[1:10, 1:6]))),
    caption = "\\label{tab:ep-2015-unprocessed}Visualization of the seven first episodes of the record number `10041`.") |>
  kable_styling(full_width = FALSE, 
                #latex_options=c("striped", "scale_down"), 
                latex_options=c("scale_down"),
                 font_size = 8) |>
  footnote(general = c("Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT_EPI: episode's weight. TUI_01: activity code. DURATION: episode's duration. LOCATION: episode's location."), threeparttable = T)

@@

The documentation of the package can be accessed by running `?ActiveCA` in `R`. The documentation includes a description of the package, information about the authors, contributors, and maintainers, and an overview of all pre-processed data sets included in the package.

Figure \ref{fig:walking_documentation} displays the documentation for the walking episodes from the 2015 GSS (`walking_2015.rda`). The documentation provides a brief description of the dataset, instructions on how to load the package, as well as details about the file format and descriptions of all variables (already labeled).

@@

For instance, to calculate the percentage of respondents from the 2015 GSS survey with active travel episodes, it is necessary to account for the person's weight. In 2015, the weight variable is represented by `WGHT_PER.` The code below demonstrates how to obtain the percentage of people with active travel episodes by age group. It uses the `dplyr` package [@dplyr2023] to manipulate the data.

The process begins by creating a dataset that sums the population by age group. Then, it joins the 2015 episodes with the 2015 Main File. Note that in both operations, the code sums the person weight variable (`WGHT_PER`) to obtain the correct population values. After this, a new dataset, `Active_percentage`, is created by merging both previous datasets. The percentage of active travel episodes by age group is calculated by dividing the total population by the population with active trip episodes, then multiplying by 100 and rounding with 2 decimal places.

@@

Table \ref{tab:main-2015-processed} shows the first ten rows and first six variables of the GSS PUMF 2015 Main File (Cycle 29), displayed in \ref{tab:main-2015-unprocessed} before our processing. Table \ref{tab:ep-2015-processed} presents the walking episodes for the record identification number `10041` from the GSS PUMF 2015 Episode File (Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}. Only the unique active travel episode appear in Table \ref{tab:ep-2015-processed} since the records were filtered to select cases with walking or cycling episodes. For both cases, Tables \ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain labeled variables, facilitating the interpretation of the data.

@@

Table \ref{tab:table-02} and \ref{tab:table-03} provide descriptive
statistics for the two modes of transportation, split by destination
categories, from 1986 to 1998 and from 2005 to 2015, respectively. In
Table \ref{tab:table-02}, one can observed that in 1986 and 1992,
walking trips destined for `home` had the highest medians. However, by
1998, the highest medians shifted to trips to `work or school`, a
transition that also occurred for cycling trips between 1992 and 1998.
Table \ref{tab:table-03} indicates that the median duration for walking trips to
`home` and `work or school` remained at 10 minutes.

@@

In 1992, walking trips with `home` as both the origin and destination
made up the majority, accounting for about 30% of all walking trips.
These trips often involved leisure activities, like short walks or dog
walking. Following this, trips from `home` to `work or school` comprised
18% of walking trips. Overall, `home` emerged as a crucial hub, either
as an origin or destination, with only 5% of trips not involving `home.`
By 2015, `home` remained a significant node, but new locations
distributed the proportion of trips to areas not considered in 1992. In
2015, the highest proportion of trips were from `home` to
`work or school` (12%) and vice versa (11%). `home` to `home` accounted
for 8% of trips, and `grocery stores` became a notable destination for
those leaving `home` (6%), surpassing trips to `other's home` (4%).

@@

For cycling trips, Figure \ref{fig:figure-02}, shows that in 1992, when
this mode of transportation was first included as an activity, the
majority of trips were from `home` to `work or school`, accounting for
about 25% of cases. This pattern remained in 2015, with these trips
representing 30% of the cases. However, a notable change occurred in
`home` to `home` trips, which decreased significantly from 19% in 1992
to 5% in 2015.

@@

This paper presents {ActiveCA}, an open data product that provides analysis-ready 
data from Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015) 
of the GSS surveys on active travel in Canada. In the form of an `R` data package, 
{ActiveCA} was developed after collecting, cleaning, and processing the survey data, 
providing information on origins, destinations, and duration of active travel, as
well other information.

It is important to remark that the present version of {ActiveCA} covers all Canadian
time use surveys up to 2015. While the most recent time use survey was carried out
in 2022 [@wray2024], the Public Use Microdata Files are currently unavailable, and
at the moment it is estimated that they will only be published in the later part
of 2025. The `R` package will be updated once these files become available.

The value of {ActiveCA} lies in its transparency, accessibility, and ease of use,
which facilitates the addition of complementary data sets in the future. `R` users
can seamlessly explore GSS walking and cycling episodes, 
with the option to suggest enhancements to the package as needed. This article adopts
the structure proposed by Anastasia and Páez [-@soukhov2023], whose work provided
essential guidance for the creation of this package. Similarly, we aim to contribute
to the academic community by promoting transparent research practices that encourage
replication and innovation in related fields. We believe that {ActiveCA} will serve
as a basis for further research on GSS and for the integration of additional data by
the authors or the wider open source community.
